Tracing the cultural evolution of meaning at the semantics-pragmatics interface

### 2016-11-18 paper/paper-v2.tex
paper revision

We distinguish between two kinds of linguistic behavior. {\em Literal interlocutors} produce
and interpret messages literally, being guided only by their lexica. {\em Pragmatic
  interlocutors} instead engage in mutual reasoning to inform their choices. Following models
of rational language use such as Rational Speech Act models \citep{frank+goodman:2012} and
their game-theoretic predecessors, the Iterated Best/Quantal Response models
\citep{franke:2009,franke+jaeger:2014}, these behaviors are captured by a reasoning
hierarchy. The hierarchy's bottom, level $0$, corresponds to literal language use. Pragmatic
language users of level $n + 1$ behave rationally according to expected level $n$ behavior of
their interlocutors. (\ref{h:level0}) and (\ref{h:leveln}) specify the behavior of literal and
pragmatic hearers of a language $L$. Their speaker counterparts are given in (\ref{s:level0})
and (\ref{s:leveln}).

\begin{flalign}
&H_{0}(s|m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha) \label{s:level0}\\
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)^\alpha) \label{s:leveln}
\end{flalign}

According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform across hearers. The behavior of literal
speakers, given in (\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the
sensitivity to which speakers prefer one signal over another based on its expected
communicative success.

>>>>

We distinguish between two kinds of pragmatic behavior. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{franke:2009,frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use. Pragmatic language users of level $n + 1$
act (approximately) rational with respect to level $n$ behavior of their
interlocutors. (\ref{h:level0}) and (\ref{s:level0}) define probabilistic behavior of literal
hearers and speakers respectively: \mf{is it a problem that $S$ denotes speakers and the set of
states?}
\begin{flalign}
&H_{0}(s|m;L) \propto pr(s) L_{sm} \label{h:level0}\\
&S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha) \label{s:level0}
\end{flalign}
According to (\ref{h:level0}), a literal hearer's interpretation of a message $m$ as a state
$s$ depends on her lexicon and her prior over states, $pr \in \Delta(S)$. For simplicity, in
the following this prior is assumed to be uniform. A literal interpreter with lexicon $L_4$
from above would then assign $\ssome$ a probability of one after hearing $\msome$, while a
literal interpreter with lexicon $L_5$ would assign $\ssome$ probability $.5$. The simplest
definition of a literal speaker would just be $S_{0}(m|s;L) \propto \exp((L_{sm})^\alpha)$, but
the more complex formulation in (\ref{s:level0})

\begin{itemize}
\item no prob zero
\item $\alpha$ adjusts the level of avoidance of untrue signaling
\end{itemize}

 The behavior of literal speakers, given in
(\ref{s:level0}), is regulated by a parameter $\alpha$ which controls the sensitivity to which
speakers prefer one signal over another based on its expected communicative success.

\begin{flalign}
&H_{n+1}(s|m;L) \propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) \propto  \exp(\lambda \; H_{n}(s|m;L)^\alpha) \label{s:leveln}
\end{flalign}
